Exoplanet population inference and the abundance \\ of Earth analogs from noisy, incomplete catalogs

### 2014-08-28 document/ms.tex
typos

In the next Section, we summarize the likelihood method for exoplanet
population inference and in \sect{hier}, we describe how to include the
effects of observational uncertainties.
The technical term for this procedure is \emph{hierarchical inference} and
while a general discussion of this field is beyond the scope of this \paper,
in the \sect{hier}, we present the basic probabilistic question and derive a
computationally tractable inference procedure.
In \sect{model}, we summarize the technique and derive the key equation for
our method: \eq{money}.
We test our method on synthetic catalogs in Sections~\sectalt{valid}
and~\sectalt{extrap}.
In \sect{real}, we use the catalog of planet candidates and the empirically
determined detection efficiency from \citet{petigura} to measure the
occurrence rate density of small planets with long orbital periods.

@@

Finally, for all the results in this \paper, we'll model the rate density as
a piecewise constant step function
\begin{eqnarray}\eqlabel{rate-model}
\rate_\ratepars (\entry) &=& \left\{\begin{array}{ll}
\exp(\ratepar_1) & \entry \in \binarea_1,\\
\exp(\ratepar_2) & \entry \in \binarea_2,\\
\cdots \\
\exp(\ratepar_J) & \entry \in \binarea_J,\\
0 & \mathrm{otherwise}
\end{array}\right.
\end{eqnarray}
where the parameters $\ratepar_j$ are the log step heights and the bins
$\binarea_j$ are fixed \foreign{a priori}.

In \app{inv-det-eff}, we use the model in \eq{rate-model} and neglect
observational uncertainties, we can derive the analytic maximum likelihood
solution for the step heights.
This result is similar to and just as simple as the
inverse-detection-efficiency method and it is guaranteed to provide a lower
variance estimate of the rate density than the standard procedure.

@@

In other words, we treat the catalog as being a dimensionality reduction of
the raw data with all the relevant information retained.
In the context of \kepler, the catalog reduces the set of downloaded pixel
time series (approximately 70,000 data points for the typical \kepler\ target)
to probabilistic constraints on a handful of physical parameters---\entry\
from above---like the orbital period and planetary radius.
If we take this set of parameters $\{\entry_k\}$ as \emph{sufficient
statistics} of the data then we can, in theory, compute \eq{crazylike}---up to
an unimportant constant---without ever looking at the raw data again!
This is important because the high-dimensional integral in \eq{crazylike}
won't generally have an analytic solution and each evaluation of the
per-object likelihood $p(\data_k\,|\,\entry_k)$ is expensive, making numerical
methods intractable.

@@

@@

We weaken the assumptions necessary for extrapolation by only assuming that
the distribution is smooth using the Gaussian process regularization described
in \sect{model}.
Under this model, the occurrence rate density at periods and radii where no
objects have been detected will be constrained---with large uncertainty---by
the heights of nearby bins.
Therefore, even though there are no candidates that qualify as Earth analogs,
we simply fit our model of the occurrence rate density in a large enough
region of parameter space (including Earth) and compute the posterior
constraints on \gammaearth.
This works because the Gaussian process regularization actually captures our
prior beliefs about the shape of the rate density function.
This model---and any other extrapolation---will, of course, break down if
there is an unmeasured sharp feature in the occurrence rate density near the
location of Earth and our method is the most conservative extrapolation
technique published to date.

For comparison, we also implemented and applied the extrapolation technique
applied by \citet{petigura}.
Their method assumes that, for small planets ($1 \le \radius/\radius_\oplus <
2$) on long periods ($\period > 50\,\mathrm{days}$), the occurrence rate
density is a flat function of logarithmic period or, equivalently, the
cumulative rate is linear.
\citet{petigura} used the candidates in their catalog to estimate the slope of
the empirical cumulative period distribution and used that function to
extrapolate.
Instead of defining \gammaearth\ differentially, as we did in \eq{gammaearth},
\citet{petigura} constrained the integral of the rate density over a box in
period and radius ($1 \le \radius/\radius_\oplus < 2$ and $200 \le
\period/\mathrm{day} < 400$).
Since their model implicitly assumes a constant rate density across the bin,
the differential rate is just their number divided by the bin area.
This rate density (rate divided by bin volume) is what is shown as a
comparison to our results in the figures.

@@

The period distribution integrated over various radius ranges is shown in
\fig{period}.
In agreement with \citet{dong}, we find that the period distribution of large
planets ($R > 8\,R_\oplus$) is inconsistent with the distribution of smaller
planets.
The rate density of large planets appears to monotonically increase as a
function of log period while the distribution for small planets seems to turn
over at a relatively short period (around 50 days) and decrease for larger
periods.

@@

\section{Comparison with previous work}

Our inferred rate density of Earth analogs (\eqalt{ge-result}) is not
consistent with previously published results.
In particular, our result is completely inconsistent with the earlier result
based on \emph{exactly the same dataset} (\citealt{petigura}).
This inconsistency is due to the different assumptions made the detailed cause
merits some investigation.
The two key differences between our analysis and previous work are \emph{(a)}
the form of the extrapolation function and \emph{(b)} the presence of
measurement uncertainties on the planet radii.

@@

For comparison, we have also included the value of \gammaearth\ implied by
\citet[][their Table 2]{dong}.
This result is based on a power law fit to the period distribution of small
planets ($R=1-2\,R_\oplus$) on long periods ($P=10-250\,\mathrm{d}$) in a
different catalog (\citealt{kepler-catalog}) with a parametric completeness
model.
There are a few factors to consider when comparing to this to our analysis.
Firstly, while \citet{dong} did use a somewhat flexible extrapolation
function than (a power law in log period), it is still very restrictive
considering the large range of periods; a broken power law might have been
more applicable.
Furthermore, their analysis did not incorporate the effects of measurement
uncertainties.
Finally, unlike the \citet{petigura}, the \citet{kepler-catalog} catalog used
by \citet{dong} includes multiple transiting systems.
As mentioned previously, the effect of this selection is hard to determine
without further investigation but it should, intuitively, cause any inference
based on the \citet{petigura} sample an underestimate of the \True\ rate.

@@

Our extrapolation of the rate density to the location of Earth is more general
and conservative than any previously published method.
We find a rate density of Earth analogs that is inconsistent with the result
published by \citet{petigura}.
This discrepancy can be largely attributed to the rigidity of the assumptions
about the period distribution.
Our extrapolation is also less confident than previous measurements.
Again, this difference is due to the fact that we allow a much more flexible
extrapolation function.
This is another illustration that, against the standard data analysis
folklore, the correct use of flexible models is \emph{conservative}.

>>>>

In the next Section, we summarize the likelihood method for exoplanet
population inference and in \sect{hier}, we describe how to include the
effects of observational uncertainties.
The technical term for this procedure is \emph{hierarchical inference} and
while a general discussion of this field is beyond the scope of this \paper,
in \sect{hier}, we present the basic probabilistic question and derive a
computationally tractable inference procedure.
In \sect{model}, we summarize the technique and derive the key equation for
our method: \eq{money}.
We test our method on synthetic catalogs in Sections~\sectalt{valid}
and~\sectalt{extrap}.
In \sect{real}, we use the catalog of planet candidates and the empirically
determined detection efficiency from \citet{petigura} to measure the
occurrence rate density of small planets with long orbital periods.

Sections~\sectalt{likelihood} and~\sectalt{hier} provide a general pedagogical
introduction to the methods used in this \paper.
Readers looking to \emph{implement} a population inference are directed to
\app{inv-det-eff} if measurement uncertainties are negligible or \sect{model}
(especially \eqalt{money}) for problems with non-negligible uncertainties.
Readers interested in our results---the inferred population of exoplanets and
Earth-analogs---can safely skip to \sect{real} and continue to the discussion
in \sect{comparison}.

@@

Finally, we model the rate density as a piecewise constant step function
\begin{eqnarray}\eqlabel{rate-model}
\rate_\ratepars (\entry) &=& \left\{\begin{array}{ll}
\exp(\ratepar_1) & \entry \in \binarea_1,\\
\exp(\ratepar_2) & \entry \in \binarea_2,\\
\cdots \\
\exp(\ratepar_J) & \entry \in \binarea_J,\\
0 & \mathrm{otherwise}
\end{array}\right.
\end{eqnarray}
where the parameters $\ratepar_j$ are the log step heights and the bins
$\binarea_j$ are fixed \foreign{a priori}.
In \app{inv-det-eff}, we use this parameterization and derive the analytic
maximum likelihood solution for the step heights.
This result is similar to and just as simple as the
inverse-detection-efficiency method and it is guaranteed to provide a lower
variance estimate of the rate density than the standard procedure.

@@

In other words, we treat the catalog as being a dimensionality reduction of
the raw data with all the relevant information retained.
In the context of \kepler, the catalog reduces the set of downloaded time
series (approximately 70,000 data points for the typical \kepler\ target) to
probabilistic constraints on a handful of physical parameters---\entry\ from
above---like the orbital period and planetary radius.
If we take this set of parameters $\{\entry_k\}$ as \emph{sufficient
statistics} of the data then we can, in theory, compute \eq{crazylike}---up to
an unimportant constant---without ever looking at the raw data again!
This is important because the high-dimensional integral in \eq{crazylike}
won't generally have an analytic solution and each evaluation of the
per-object likelihood $p(\data_k\,|\,\entry_k)$ is expensive, making numerical
methods intractable.

@@

\dfmtodo{Add a few more pedagogical words here.}

@@

We weaken the assumptions necessary for extrapolation by only assuming that
the distribution is smooth using the Gaussian process regularization described
in \sect{model}.
Under this model, the occurrence rate density at periods and radii where no
objects have been detected will be constrained---with large uncertainty---by
the heights of nearby bins.
Therefore, even though there are no candidates that qualify as Earth analogs,
we simply fit our model of the occurrence rate density in a large enough
region of parameter space (including Earth) and compute the posterior
constraints on \gammaearth.
This works because the Gaussian process regularization actually captures our
prior beliefs about the shape of the rate density function.
This model---and any other extrapolation---will, of course, break down if
there is an unmeasured sharp feature in the occurrence rate density near the
location of Earth but our method is the most conservative extrapolation
technique published to date.

For comparison, we also implemented and applied the extrapolation technique
applied by \citet{petigura}.
Their method assumes that, for small planets ($1 \le \radius/\radius_\oplus <
2$) on long periods ($\period > 50\,\mathrm{days}$), the occurrence rate
density is a flat function of logarithmic period or, equivalently, the
cumulative rate is linear.
\citet{petigura} used the candidates in their catalog to estimate the slope of
the empirical cumulative period distribution and used that function to
extrapolate.
Instead of defining \gammaearth\ differentially, as we did in \eq{gammaearth},
\citet{petigura} constrained the integral of the rate density over a box in
period and radius ($1 \le \radius/\radius_\oplus < 2$ and $200 \le
\period/\mathrm{day} < 400$).
Since their model implicitly assumes a constant rate density across the bin,
the differential rate is just their number divided by the bin volume.
This rate density (rate divided by bin volume) is what is shown as a
comparison to our results in the figures.

@@

The period distribution integrated over various radius ranges is shown in
\fig{period}.
In agreement with \citet{dong}, we find that the period distribution of large
planets ($R > 8\,R_\oplus$) is inconsistent with the distribution of smaller
planets.
The rate density of large planets appears to monotonically increase as a
function of log period while the distribution for small planets seems to turn
over at a relatively short period (around 50 days) and decrease for longer
periods.

@@

\section{Comparison with previous work}
\sectlabel{comparison}

Our inferred rate density of Earth analogs (\eqalt{ge-result}) is not
consistent with previously published results.
In particular, our result is completely inconsistent with the earlier result
based on \emph{exactly the same dataset} (\citealt{petigura}).
This inconsistency is due to the different assumptions made and the detailed
cause merits some investigation.
The two key differences between our analysis and previous work are \emph{(a)}
the form of the extrapolation function, and \emph{(b)} the presence of
measurement uncertainties on the planet radii.

@@

For comparison, we have also included the value of \gammaearth\ implied by
\citet[][their Table 2]{dong}.
This result is based on a power law fit to the period distribution of small
planets ($R=1-2\,R_\oplus$) on long periods ($P=10-250\,\mathrm{d}$) in a
different catalog (\citealt{kepler-catalog}) with a parametric completeness
model.
There are a few factors to consider when comparing to this to our analysis.
Firstly, while \citet{dong} fit a power law in log period, this is still a
very restrictive model when considering this large range of periods.
A broken power law might be more applicable.
Furthermore, their analysis did not incorporate the effects of measurement
uncertainties.
Finally, unlike the \citet{petigura}, the \citet{kepler-catalog} catalog used
by \citet{dong} includes multiple transiting systems.
As mentioned previously, the effect of this selection is hard to determine
without further investigation but it should, intuitively, cause any inference
based on the \citet{petigura} sample to be an underestimate of the \True\
rate.

@@

Our extrapolation of the rate density to the location of Earth is more general
and conservative than any previously published method.
We find a rate density of Earth analogs that is inconsistent with the result
published by \citet{petigura}.
This discrepancy can be attributed to both the rigidity of the assumptions
about the period distribution and the effects of non-negligible measurement
uncertainties.
Our extrapolation is also less confident than previous measurements.
Again, this difference is due to the fact that we allow a much more flexible
extrapolation function.
This is another illustration that, against the standard data analysis
folklore, the correct use of flexible models is \emph{conservative}.
